\TitlePt

### 2016-03-01 paper1.tex
Include note about being published in fatiando chapter
Add references to Fatiando paper

\chapter{Modeling the Earth with Fatiando a Terra}\label{chap:fatiando}

@@

The seismic method is among the most widely studied due to the high
industry demand. Thus, a range of well established open-source software
have been developed for seismic processing. These include
\href{http://www.cwp.mines.edu/cwpcodes/}{Seismic Un*x} (SU) {[}SU{]}\_,
\href{http://www.ahay.org/}{Madagascar} {[}MAD{]}\_,
\href{http://opendtect.org}{OpendTect}, and
\href{http://www.gebrproject.com}{GêBR}. A noteworthy open-source
project that is not seismic related is the
\href{http://gmt.soest.hawaii.edu/}{Generic Mapping Tools} (GMT) project
{[}GMT{]}\_. The GMT are a well established collection of command-line
programs for plotting maps with a variety of different map projections.
For geodynamic modeling there is the
\href{http://www.geodynamics.org}{Computational Infrastructure for
Geodynamics} (CIG), which has grouped various well documented software
packages. However, even with this wide range of well maintained software
projects, many geophysical modeling software that are provided online
still have no open-source license statement, have cryptic I/O files, are
hard to integrate into a pipeline, and make code reuse and remixing
challenging. Some of these problems are being worked on by the
\href{http://geosys.usc.edu/projects/seatree/}{Solid Earth Teaching and
Research Environment} (SEATREE) {[}SEATREE{]}\_ by providing a common
graphical interface for previously existing software. The numerical
computations are performed by the pre-existing underlying C/Fortran
programs. Conversely, the SEATREE code (written in Python) handles the
I/O and user interface. This makes the use of these tools easier and
more approachable to students. However, the lack of a common API means
that the code for these programs cannot be easily combined to create new
modeling tools.

\href{http://www.fatiando.org}{Fatiando a Terra} aims at providing such
an API for geophysical modeling. Functions in the \texttt{fatiando}
package use compatible data and mesh formats so that the output of one
modeling function can be used as input for another. Furthermore,
routines can be combined and reused to create new modeling algorithms.
Fatiando a Terra also automates common tasks such as griding, map
plotting with \href{http://matplotlib.org}{Matplotlib} {[}MPL{]}\_, and
3D plotting with
\href{http://code.enthought.com/projects/mayavi}{Mayavi} {[}MYV{]}\_.
Version 0.1 of Fatiando a Terra is focused on gravity and magnetic
methods because this is the main focus of the developers. However,
simple ``toy'' problems for seismology and geothermics are available and
can be useful for teaching geophysics.

The following sections illustrate the functionality and design of
Fatiando a Terra using various code samples. An
\href{http://ipython.org/}{IPython} {[}IPY{]}\_ notebook file with these
code samples is provided by {[}SAMPLES{]}\_ at
\url{http://dx.doi.org/10.6084/m9.figshare.708390}.

@@

The \texttt{fatiando.vis.myv} module contains functions to automate 3D plotting
using Mayavi {[}MYV{]}\_. The \texttt{mayavi.mlab} interface requires geometric
elements to be formatted as TVTK objects. Thus, plotting functions in
\texttt{fatiando.vis.myv} automatically create TVTK representations of
\texttt{fatiando.mesher} objects and plot them using a suitable function of
\texttt{mayavi.mlab}. Also included are utility functions for drawing axes,
walls on the figure bounding box, etc. For example, the
\texttt{fatiando.vis.myv.figure} function creates a figure and rotates it so
that the z-axis points down, as is standard in geophysics. The following
example shows how to plot the 3D right rectangular prism model that we created
previously (Figure~\ref{fig:twoprisms}):

@@

The module \texttt{fatiando.gravmag.polyprism} implements the method of
{[}PLOUFF{]}\_ to forward model the gravity fields of a 3D right
polygonal prism. The following code sample shows how to interactively
generate a polygonal prism model and calculate its gravity anomaly
(Figures~\ref{fig:drawing} and \ref{fig:polyprism}):

@@

The module \texttt{fatiando.gravmag.imaging} implements the imaging
methods described in {[}FP{]}\_. These methods aim to produce an image
of the geologic source from the observed gravity or magnetic data. The
following code sample uses the ``sandwich model'' method {[}SNDW{]}\_ to
image the polygonal prism, produced in the previous section, based on
its gravity anomaly (Figure~\ref{fig:imaging}):

@@

Also implemented in Fatiando a Terra are some recent developments in
gravity and magnetic inversion methods. The method of ``planting
anomalous densities'' by {[}UB{]}\_ is implemented in the
\texttt{fatiando.gravmag.harvester} module. In contrast to imaging
methods, this is an inversion method, i.e., it estimates a physical
property distribution (density in the case of gravity data) that fits
the observed data. This particular method requires the user to specify a
``seed'' (Figure~\ref{fig:seed}) around which the estimated density
distribution grows (Figure~\ref{fig:harvester}):

@@

Even though the implementation in \texttt{fatiando.seismic.srtomo} is greatly
simplified and not usable in real tomography problems, the result in
Figure~\ref{fig:tomo} illustrates interesting inverse problem concepts.  Notice
how the estimated velocity is blurred in the corners where no rays pass
through. This is because the data (travel-times) provide no information about
the velocity in those areas. Areas like those constitute the null space of the
inverse problem {[}MENKE{]}\_, where any velocity value estimated will provide
an equal fit to the data.  Thus, the tomography problem requires the use of
prior information in the form of regularization. Most commonly used in
tomography problems is the Tikhonov first-order regularization, e.g., a
smoothness constraint {[}MENKE{]}\_. The amount of smoothness imposed on the
solution is controlled by the \texttt{smooth} argument of function
\texttt{fatiando.seismic.srtomo.run}. That is how we are able to estimate a
unique and stable solution and why the result is specially smoothed where there
are no rays.

@@

Fatiando a Terra enables quick prototyping of new algorithms because of
the collection of fast forward modeling routines and the simple syntax
and high level of the Python language. After prototyping, the
performance bottlenecks of these algorithms can be easily diagnosed
using the advanced profiling tools available in the Python language.
Optimization of only small components of code can be done without loss
of flexibility using the Cython language {[}CYTHON{]}\_.

>>>>

\chapter{Modeling the Earth with Fatiando a Terra}
\label{chap:fatiando}

This chapter was published in the Proceedings of the 12th Python in Science
Conference (Scipy 2013).
\url{http://conference.scipy.org/proceedings/scipy2013/uieda.html}

@@

The seismic method is among the most widely studied due to the high
industry demand. Thus, a range of well established open-source software
have been developed for seismic processing. These include
\href{http://www.cwp.mines.edu/cwpcodes/}{Seismic Un*x} (SU)
\citep{stockwelljr.1999},
\href{http://www.ahay.org/}{Madagascar} \citep{madagascardevelopmentteam2013},
\href{http://opendtect.org}{OpendTect}, and
\href{http://www.gebrproject.com}{GêBR}. A noteworthy open-source
project that is not seismic related is the
\href{http://gmt.soest.hawaii.edu/}{Generic Mapping Tools}
(GMT) project
\citet{wessel1991}. The GMT are a well established collection of command-line
programs for plotting maps with a variety of different map projections.
For geodynamic modeling there is the
\href{http://www.geodynamics.org}{Computational Infrastructure for
Geodynamics}, which has grouped various well documented software
packages. However, even with this wide range of well maintained software
projects, many geophysical modeling software that are provided online
still have no open-source license statement, have cryptic I/O files, are
hard to integrate into a pipeline, and make code reuse and remixing
challenging. Some of these problems are being worked on by the
\href{http://geosys.usc.edu/projects/seatree/}{Solid Earth Teaching and
Research Environment} (SEATREE) \citet{milner2009} by providing a common
graphical interface for previously existing software. The numerical
computations are performed by the pre-existing underlying C/Fortran
programs. Conversely, the SEATREE code (written in Python) handles the
I/O and user interface. This makes the use of these tools easier and
more approachable to students. However, the lack of a common API means
that the code for these programs cannot be easily combined to create new
modeling tools.

\href{http://www.fatiando.org}{Fatiando a Terra} aims at providing such an API
for geophysical modeling. Functions in the \texttt{fatiando} package use
compatible data and mesh formats so that the output of one modeling function
can be used as input for another. Furthermore, routines can be combined and
reused to create new modeling algorithms.  Fatiando a Terra also automates
common tasks such as griding, map plotting with
\href{http://matplotlib.org}{Matplotlib} \citep{hunter2007}, and 3D plotting
with \href{http://code.enthought.com/projects/mayavi}{Mayavi}
\citep{ramachandran2011}.  Version 0.1 of Fatiando a Terra is focused on
gravity and magnetic methods because this is the main focus of the developers.
However, simple ``toy'' problems for seismology and geothermics are available
and can be useful for teaching geophysics.

The following sections illustrate the functionality and design of
Fatiando a Terra using various code samples. An
\href{http://ipython.org/}{IPython} \citep{perez2007} notebook file with these
code samples is provided by \citet{uieda2013} at
\url{http://dx.doi.org/10.6084/m9.figshare.708390}.

@@

The \texttt{fatiando.vis.myv} module contains functions to automate 3D plotting
using Mayavi \citep{ramachandran2011}. The \texttt{mayavi.mlab} interface
requires geometric elements to be formatted as TVTK objects. Thus, plotting
functions in \texttt{fatiando.vis.myv} automatically create TVTK
representations of \texttt{fatiando.mesher} objects and plot them using a
suitable function of \texttt{mayavi.mlab}. Also included are utility functions
for drawing axes, walls on the figure bounding box, etc. For example, the
\texttt{fatiando.vis.myv.figure} function creates a figure and rotates it so
that the z-axis points down, as is standard in geophysics. The following
example shows how to plot the 3D right rectangular prism model that we created
previously (Figure~\ref{fig:twoprisms}):

@@

The module \texttt{fatiando.gravmag.polyprism} implements the method of
\citet{plouff1976} to forward model the gravity fields of a 3D right polygonal
prism. The following code sample shows how to interactively generate a
polygonal prism model and calculate its gravity anomaly
(Figures~\ref{fig:drawing} and \ref{fig:polyprism}):

@@

The module \texttt{fatiando.gravmag.imaging} implements the imaging methods
described in \citet{fedi2012}. These methods aim to produce an image of the
geologic source from the observed gravity or magnetic data. The following code
sample uses the ``sandwich model'' method \citep{pedersen1991} to image the
polygonal prism, produced in the previous section, based on its gravity anomaly
(Figure~\ref{fig:imaging}):

@@

Also implemented in Fatiando a Terra are some recent developments in
gravity and magnetic inversion methods. The method of ``planting
anomalous densities'' by \citet{uieda2012} is implemented in the
\texttt{fatiando.gravmag.harvester} module. In contrast to imaging
methods, this is an inversion method, i.e., it estimates a physical
property distribution (density in the case of gravity data) that fits
the observed data. This particular method requires the user to specify a
``seed'' (Figure~\ref{fig:seed}) around which the estimated density
distribution grows (Figure~\ref{fig:harvester}):

@@

Even though the implementation in \texttt{fatiando.seismic.srtomo} is greatly
simplified and not usable in real tomography problems, the result in
Figure~\ref{fig:tomo} illustrates interesting inverse problem concepts.  Notice
how the estimated velocity is blurred in the corners where no rays pass
through. This is because the data (travel-times) provide no information about
the velocity in those areas. Areas like those constitute the null space of the
inverse problem \citep{menke1984}, where any velocity value estimated will
provide an equal fit to the data.  Thus, the tomography problem requires the
use of prior information in the form of regularization. Most commonly used in
tomography problems is the Tikhonov first-order regularization, e.g., a
smoothness constraint \citep{menke1984}. The amount of smoothness imposed on
the solution is controlled by the \texttt{smooth} argument of function
\texttt{fatiando.seismic.srtomo.run}. That is how we are able to estimate a
unique and stable solution and why the result is specially smoothed where there
are no rays.

@@

Fatiando a Terra enables quick prototyping of new algorithms because of
the collection of fast forward modeling routines and the simple syntax
and high level of the Python language. After prototyping, the
performance bottlenecks of these algorithms can be easily diagnosed
using the advanced profiling tools available in the Python language.
Optimization of only small components of code can be done without loss
of flexibility using the Cython language \citep{behnel2011}.
